Monte Carlo solvers for sparse linear systems

### 2016-01-25 MAX_reports/MC_solvers_article/convergence.tex

\section{Convergence Behavior of Stochastic Methods}

The requests imposed on the expected value and second moment of the estimators
can be reformulated in a purely deterministic setting by looking at the
spectral radii of some properly defined matrices.

For instance the condition of finiteness of the expected value can be
reformulated by requiring
\begin{equation}
 \rho(H)<1
 \label{rho_h},
\end{equation}

where $H$ is the iteration matrix of the fixed point scheme.
In fact both equations (\ref{dir_mean}) and (\ref{adj_mean}) can be considered
as
power series in terms of $H$ and the spectral radius of $H$ smaller
than 1 is a necessary and sufficient condition for the Neumann series to
converge.

By analogy we aim at reproducing a similar reasoning for the second moment of
the estimator, since we want it to be finite as well.
By looking at the equations of the variance (\ref{dir_var}) and (\ref{adj_var})
for
the Forward and the Adjoint method respectively, we notice that the second
moment can be reinterpreted as a power series with respect to the matrices
defined as follows:

\[
 \hat{H}_{i,j}=\frac{H_{i,j}^2}{P_{i,j}} \quad \text{- Forward Method}.
\]

or

\[
 \hat{H}_{i,j}=\frac{H_{j,i}^2}{P_{i,j}} \quad \text{- Adjoint Method}.
\]

Accordingly to the necessary and sufficient condition for the power series to
converge, we require
\begin{equation}
 \rho(\hat{H})<1.
 \label{rho_star}
\end{equation}

Condition (\ref{rho_h}) must be required for a generic fixed point scheme to
reach convergence, whereas the extra condition (\ref{rho_star}) is typical of
the stochastic schemes presented in this work. Having two conditions to
verify makes the numerical treatment of the linear system via MC very
difficult. We will discuss this in details later in the treatise. \newline

@@

Here below some results presented in
\cite{MASC2013} about necessary conditions and sufficient
conditions for
convergence are enunciated. In particular a suitable choice for constructing a
transition probability is discussed to guarantee and speed-up convergence.

We remember that the construction of the transition probability must satisfy
some constraints such as

\[
\begin{cases}
  P_{i,j}\ge 0 \\
 \sum_{j=1}^N P_{i,j}=1
\end{cases}
\]

plus one more restriction which varies depending on the particular Monte Carlo
approach chosen

\begin{align*}
& \text{Forward Method:} \quad H_{i,j}\ne 0 \Rightarrow P_{i,j}\ne 0 \\ &
\text{Adjoint Method:} \quad H_{j,i}\ne 0 \Rightarrow P_{i,j}\ne 0
\end{align*}.

The previous restrictions on the attainable values for the entries of the
transition probability are called \textit{transition conditions}.

At first an introductory theoretical result is shown, necessary to motivate the
choices made afterwards for the choice of the transition probability.

@@

It might be observed that the minimum of the quantity $\displaystyle
\sum_{k=0}^N\frac{a_k^2}{p_k}$ is attained when the probability vector is
defined such as $\displaystyle p_k=\frac{\lvert a_k\rvert}{\sum_{k=1}^N \lvert
a_k\rvert}$. It implies that the $\infty$-norm of $\hat{H}$ is minimized as
much
as possible. \newline

For the sake of simplicity let us introduce now a generic random walk
truncated at a certain $k$-th step
\[
 \gamma_k:\; r_0\rightarrow r_1 \rightarrow r_2 \rightarrow \cdots \rightarrow
r_k
\]
and the statistical estimator
\[
 X(\gamma_k)=\frac{H_{r_0,r_1}H_{r_1,r_2}\cdots
H_{r_{k-1},r_k}}{P_{r_0,r_1}P_{r_1,r_2}\cdots P_{r_{k-1},r_k}}f_{r_k}
\]
which is associated with the Forward Method.

@@

If we introduce the estimator
\[
 X(\gamma_k)=\frac{H^T_{r_0,r_1}H^T_{r_1,r_2}\cdots
H^T_{r_{k-1},r_k}}{P_{r_0,r_1}P_{r_1,r_2}\cdots
P_{r_{k-1},r_k}}sign(f_{r_0})\lVert \mathbf{f}\rVert_1
\]
which is associated with the Adjoint Method, then we can state a
theorem specular to \ref{for_thm}.

@@

In particular, relying on what guaranteed by Theorem \ref{lemma}, in both
Theorems \ref{for_thm} and \ref{adj_thm} the thesis holds by picking the MAO
transition probability.

These theoretical results represent sufficient conditions for the convergence
of the
Forward and Adjoint Monte Carlo and they can be easily checked.
However requiring $\lVert
H\rVert_{\infty}<1$ or $\lVert H\rVert_1<1$ may be too demanding.
Indeed, even if there exist some preconditioners able to guarantee this
condition (e.g. approximate inverse preconditioners with a proper value of
dropping
tolerance), it may entail to compromise the sparsity of the
iteration
matrix.

Thusly, in situations where the hypotheses of Theorems \ref{for_thm} and
\ref{adj_thm} are too demanding to be verified, the necessary and sufficient
condition
on the
spectral radius of $\hat{H}$ is the only criterion to take as reference.
Nonetheless this constraint consists in
computing the spectral radii of two matrices and the calculation
might be very costly. Moreover the computation of the spectral radii is
affected by resilience issues as well as
a deterministic solver for the resolution of
a linear system. Therefore the original difficulty is transferred without being
actually tackled.

\subsubsection{Forward MC}

\subsubsection{Adjoint MC with Collision Estimator}

\subsubsection{Adjoint MC with Expected-Value Estimator}

@@

The way the transition probability is defined is extremely important. In fact
in some circumstances it might determine whether the stochastic scheme
converges or not. Two very simple ways of defining it are listed here
below. These are the approaches adopted typically in literature.

\subsubsection{Uniform probabilities}
With this approach transition matrix $P$ is such that all the non-zero elements
in each row have equal probability of occurring.
\[
\text{Forward}:\;P_{i,j}=
\begin{cases}
0 \quad \quad \quad \qquad \qquad \qquad \text{if}\quad H_{i,j}=0 \\
\frac{1}{\#(\text{non-zeros in the row)}} \quad \text{if} \quad H_{i,j}\ne 0
\end{cases}\quad
\]
\[
\text{Adjoint}:\;P_{i,j}=
\begin{cases}
0 \quad \quad \quad \qquad \qquad \qquad \text{if}\quad H_{j,i}=0 \\
\frac{1}{\#(\text{non-zeros in the column)}} \quad \text{if} \quad H_{j,i}\ne 0
\end{cases}
\]
The Monte Carlo approach resorting to this definition of the transition matrix,
in accordance to \cite{AADBTW2005}, is called \textit{Uniform Monte Carlo} (UM).

\subsubsection{Weighted probabilities}
An alternative way of defining the transition matrix consists in attributing
probabilities to all the entries of the transition probability accordingly
to the magnitude of the elements. For instance we may employ the following
definition
\[
\text{Forward}: \;
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert
H_{i,j}\rvert^p}{\sum_{k=1}^n
\lvert H_{i,k}\rvert^p}
\]

\[
\text{Adjoint}: \;
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert
H_{j,i}\rvert^p}{\sum_{k=1}^n
\lvert H_{k,i}\rvert^p}
\]

where $p\in \mathbb{N}$.
This way of proceeding, for $p=1$, still basing on the work \cite{AADBTW2005},
is called
\textit{Monte Carlo Almost Optimal} (MAO).
The reason why the almost optimal probability is called this way is associated
with a notice from Theorem \ref{lemma}. Indeed from
the mentioned lemma we know that the quantity $\displaystyle
\sum_{k=0}^N\frac{a_k^2}{p_k}$ is minimized when the probability vector is
defined such as $\displaystyle p_k=\frac{\lvert a_k\rvert}{\sum_{k=1}^N \lvert
a_k\rvert}$. This implies that the almost optimal probability minimizes the
$\infty$-norm of $\hat{H}$ in the Forward method and the $1$-norm of $\hat{H}$
in the
Adjoint method. This is very important to be pointed out since these two norms
are so far the best practical upper bounds for $\rho(\hat{H})$ that we can
easily
handle. In the Section dedicated to the numerical results it will be proved
that the Almost Optimal Probability eases the convergence of the MC solver,
even in cases when the employment of the Uniform Probability does not provide
convergence at all.

@@

As we have already seen, sufficient conditions for
the convergence of the Monte Carlo linear solvers are very restrictive (see
\cite{MASC2013}), whereas the necessary and sufficient condition requires the
computation of
$\rho(H)$and $\rho(\hat{H})$, which is very expensive in general. Once a
proper
preconditioner is picked (e.g. Approximate Inverse), $\rho(H)<1$ may
be always guaranteed, whilst $\rho(\hat{H})<1$ is more
problematic to be verified.

In order to bypass the cost of these checks, it is useful to pursue for sets of
matrices for
which the Monte Carlo linear solver always converges, maybe with the use of an
ad-hoc preconditioner.

\subsubsection{Strictly Diagonally Dominant}

@@

Introducing a MAO transition probability for the Forward Method we get
\[
 P_{ij}=\frac{\lvert H_{ij}\rvert}{\sum_{k=1}^n\lvert H_{ik}\rvert}
\]
therefore the entries of $\hat{H}$ are defined as follows
\[
 \hat{H}_{ij}=\frac{H^2_{ij}}{P_{ij}}=\lvert H_{ij}\rvert\bigg
(\sum_{k=1}^n\lvert
H_{ik}\rvert\bigg ).
\]
Therefore
\[
 \sum_{j=1}^n\lvert \hat{H}_{ij} \rvert = \sum_{j=1}^n \hat{H}_{ij} =\bigg
(\sum_{j=1}^n \lvert H_{ij}\rvert\bigg )\bigg (\sum_{k=1}^n\lvert
H_{ik}\rvert\bigg ) =\bigg
(\sum_{j=1}^n \lvert H_{ij}\rvert\bigg ) ^2 <1 \quad \forall i=1,\cdots, n.
\]
This automatically implies that $\rho(\hat{H})<=\lVert
\hat{H}\rVert_{\infty}<1$. Thus the Forward Monte Carlo converges.
However nothing a priori can be
stated about the behavior of the Adjoint method.\newline

Instead if (\ref{sddc}) holds we can resort to a right diagonal preconditioning
and we
get an iteration matrix $H=I-AD^{-1}$ such that $\lVert H \rVert_{1}<1$.
In this case, by following a similar reasoning to the one made before, we
conclude that the Adjoint Method convergences thanks to $\lVert
\hat{H}\rVert_1<1$,
whereas nothing can be said in advance as concerns the Forward method.

\subsubsection{Generalized Diagonally Dominant}

@@

A proper subset of the generalized diagonally dominant matrices is represented
by $M$- matrices (see \cite{Ax1996}).
There are many definitions of $M$-matrices that may be found on books and all
of them are equivalent. The one presented in \cite{Saad} states the
following

\begin{defn}
A matrix $A\in\mathbb{R}^{n\times n}$ is said to be an $M$-matrix if it
satisfies the following four properties:
\begin{itemize}
 \item $a_{i,i}>0$ for $i=1,\cdots,n$
 \item $a_{i,j}\le 0$ for $i\ne j$, $i,j=1,\cdots,n$
 \item $A$ is nonsingular
 \item $A^{-1}\ge 0$
\end{itemize}
\end{defn}

A very important result about $M$-matrices enables to come up with a convergent
splitting and, therefore, a convergent fixed point scheme.

In particular Jacobi, Block Jacobi and Gauss Seidel  are
convergent
splitting for an $M$-matrix.
However, the construction of a splitting like this is not enough to ensure the
convergence of
the Monte Carlo linear solver. In fact none of these conditions is sufficient
to guarantee $\rho(\hat{H})<1$. Because of this obstacle, other options must
be considered.\newline

@@

Here below we report the algorithm at hand\newline

@@

This approach turns a generalized diagonally dominant matrix into a
strictly diagonally dominant matrix by rows. By substituting
$\displaystyle S_{i}=\sum^{n}_{\substack{j=1 \\ j\ne i}}\lvert a_{ij}\rvert$
at step 1 with
$\displaystyle S_{i}=\sum^{n}_{\substack{i=1 \\ i\ne j}}\lvert a_{ij}\rvert$
and by replacing
$a_{ji}=a_{ji}\cdot d_i$ with $a_{ji}=a_{ji}\cdot d_j$ we obtain the algorithm
that turns a GDDM matrix into a s.d.d. by columns.\newline

@@

\subsubsection{Block Diagonally Dominant}

 In this section we wonder in which situations a block diagonal preconditioning
might come up with a convergent Monte Carlo linear solver.
By mimicking the computations we already showed previously, the iteration
matrix $H\in\mathbb{R}^{n\times n}$ resulting from a block diagonal
preconditioning takes the form

\[
 H=I-D^{-1}A=\begin{bmatrix}0_{n_1\times n_1} & -[A_{11}]^{-1}A_{12} & \cdots &
\cdots & -[A_{11}]^{-1}A_{1p} \\
-[A_{22}]^{-1}A_{21} & 0_{n_2\times n_2} & -[A_{22}]^{-1}A_{23} &
\cdots & -[A_{22}]^{-1}A_{2p}\\
\vdots & \vdots & \ddots & \vdots & \vdots\\
\vdots & \vdots & \vdots &\ddots & \vdots \\
-[A_{pp}]^{-1}A_{p1} &  \cdots & \cdots&
-[A_{pp}]^{-1}A_{p,p-1} & 0_{n_p \times n_p}
\end{bmatrix}.
\]

For the sake of clarity, we notify that the symbol "$\%$" used in the following
subsections represents the modulus function between two integer numbers, while
the floor function is represented with pretty standard symbol "$\lfloor
\cdot \rfloor$".

\paragraph{Forward method}
By assuming that all $n_i$'s have the same value we may define
\[
 m=n_i=\text{size of a block}=\frac{n}{p}.
\]

The MAO transition probability matrix becomes:
\[
 P_{i,j}=\frac{\lvert H_{i,j} \rvert}{\sum_{k=1}^n\lvert H_{i,k}
\rvert}=\frac{\bigg \lvert \bigg ([A_{\lfloor
\frac{i}{m}\rfloor \lfloor]
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg
\rvert}{\sum_{\substack{k=1\\k\ne i}}^n\bigg \lvert \bigg
([A_{\lfloor
\frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{k}{m}\rfloor}\bigg )_{i\%m,k\%m}\bigg \rvert}
\]

Consequently, the $\hat{H}$ matrix is defined such that
\[
\hat{H}_{i,j} = \lvert H_{i,j}\rvert\bigg(\sum_{k=1}^n\lvert H_{ik}\rvert\bigg)=
\bigg \lvert \bigg ([A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg
\rvert
\sum_{\substack{k=1\\k\ne i}}^n\bigg \lvert \bigg ([A_{\lfloor
\frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{k}{m}\rfloor}\bigg )_{i\%m,k\%m}\bigg \rvert
\]

By computing the sum over a generic row of $\hat{H}$ we get:
\[
 \sum_{j=1}^n \lvert \hat{H}_{i,j}\rvert=\sum_{j=1}^n \hat{H}_{i,j} =
 \bigg ( \sum_{\substack{j=1\\j\ne i}}^n\bigg \lvert \bigg ([A_{\lfloor
\frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg \rvert \bigg ) ^2.
\]

If we focus on the norm $\lVert \hat{H}\rVert_{\infty}$, then the following
equivalence condition holds:
\[
 \lVert \hat{H}\rVert_{\infty}<1 \Leftrightarrow \sum_{\substack{j=1\\j\ne
i}}^n\bigg \lvert \bigg ([A_{\lfloor
\frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg \rvert <1 \quad \forall
i=1,\cdots,n
\]
 A sufficient condition for this to happen is
 \begin{equation}
  \sum_{\substack{j=1\\j\ne i}}^p \lVert [A_{ii}]^{-1}A_{ij}\rVert_{\infty}<1.
    \label{block_cs}\quad \forall i=1,\ldots,p.
 \end{equation}

By defining a matrix $\tilde{H}\in \mathbb{R}^{p\times p}$ such that
\[
 \tilde{H}=\begin{bmatrix}0 & \lVert [A_{11}]^{-1}A_{12}\rVert_{\infty} &
\cdots &
\cdots & \lVert [A_{11}]^{-1}A_{1p}\rVert_{\infty} \\
\lVert [A_{22}]^{-1}A_{21}\rVert_{\infty} & 0 & \lVert
[A_{22}]^{-1}A_{23}\rVert_{\infty} &
\cdots & \lVert [A_{22}]^{-1}A_{2p}\rVert_{\infty} \\
\vdots & \vdots & \ddots & \vdots & \vdots\\
\vdots & \vdots & \vdots &\ddots & \vdots \\
\lVert [A_{pp}]^{-1}A_{p1}\rVert_{\infty} &  \cdots & \cdots&
\lVert [A_{pp}]^{-1}A_{p,p-1}\rVert_{\infty} & 0
\end{bmatrix}
\]

We can use the $\tilde{H}$ matrix just defined in order to introduce a
sufficient condition for the convergence of the Forward Monet Carlo method with
a Block Diagonal preconditioning.

\begin{equation}
 \lVert \tilde{H} \rVert_{\infty}<1 \Rightarrow \lVert \hat{H}
\rVert_{\infty}<1.
\end{equation}

>>>>

\section{Convergence Behavior of Stochastic Methods}
\label{sec:convergence}

The convergence requirements imposed by the Monte Carlo estimator and
the corresponding variance can be reformulated in a purely deterministic
setting by considering certain properly defined matrices.
For instance, the condition of finiteness of the expected value can be
reformulated by requiring
\begin{equation}
 \rho(H)<1
 \label{rho_h},
\end{equation}
where $H$ is the iteration matrix of the fixed point scheme.
In fact both equations (\ref{dir_mean}) and (\ref{adj_mean}) can be considered
as
power series in terms of $H$ and the spectral radius of $H$ smaller
than 1 is a necessary and sufficient condition for the Neumann series to
converge.

We wish to achieve a similar result based on the requirement that the
second moment of the estimator must be finite as well.
Examination of the equations of the variance (\ref{dir_var}) and (\ref{adj_var})
for
the forward and the adjoint method respectively, we notice that the second
moment can be reinterpreted as a power series with respect to the matrices
defined as follows:
\[
 \hat{H}_{i,j}=\frac{H_{i,j}^2}{P_{i,j}} \quad \text{- Forward Method}.
\]
or
\[
 \hat{H}_{i,j}=\frac{H_{j,i}^2}{P_{i,j}} \quad \text{- Adjoint Method}.
\]
Accordingly to the necessary and sufficient condition for the power series to
converge, we require
\begin{equation}
 \rho(\hat{H})<1.
 \label{rho_star}
\end{equation}
Condition (\ref{rho_h}) must be required for a generic fixed point scheme to
reach convergence, whereas the extra condition (\ref{rho_star}) is typical of
the stochastic schemes presented in this work.

@@

Here we briefly reproduce some results introduced in \cite{MASC2013}
concerning necessary conditions and sufficient conditions for convergence.
In particular, a suitable choice for constructing a
transition probability is discussed to guarantee convergence.

The construction of the transition probability must satisfy
\[
\begin{cases}
  P_{i,j}\ge 0 \\
 \sum_{j=1}^N P_{i,j}\le 1 \,.
\end{cases}
\]
One additional constraint relates the sparsity pattern of $H$ to that
of the transition probabilities:
\begin{align*}
& \text{Forward Method:} \quad H_{i,j}\ne 0 \Rightarrow P_{i,j}\ne 0 \\ &
\text{Adjoint Method:} \quad H_{j,i}\ne 0 \Rightarrow P_{i,j}\ne 0
\end{align*}.
The previous constraints on the values for the entries of the
transition probability are called \textit{transition conditions}.

An introductory theoretical result is provided to motivate the possible
choices for the transition probability.

@@

It might be observed that the minimum of the quantity $\displaystyle
\sum_{k=0}^N\frac{a_k^2}{p_k}$ is attained when the probability vector is
defined such as $\displaystyle p_k=\frac{\lvert a_k\rvert}{\sum_{k=1}^N \lvert
a_k\rvert}$. It implies that the $\infty$-norm of $\hat{H}$ is minimized as
much
as possible.

For the sake of simplicity let us introduce now a generic random walk
truncated at a certain $k$-th step
\[
 \gamma_k:\; r_0\rightarrow r_1 \rightarrow r_2 \rightarrow \cdots \rightarrow
r_k
\]
and the statistical estimator
\[
 X(\gamma_k)=\frac{H_{r_0,r_1}H_{r_1,r_2}\cdots
H_{r_{k-1},r_k}}{P_{r_0,r_1}P_{r_1,r_2}\cdots P_{r_{k-1},r_k}}f_{r_k}
\]
which is associated with the forward method.

@@

If we introduce the estimator
\[
 X(\gamma_k)=\frac{H^T_{r_0,r_1}H^T_{r_1,r_2}\cdots
H^T_{r_{k-1},r_k}}{P_{r_0,r_1}P_{r_1,r_2}\cdots
P_{r_{k-1},r_k}}sign(f_{r_0})\lVert \mathbf{f}\rVert_1
\]
which is associated with the adjoint method, then we can state a
theorem specular to \ref{for_thm}.

@@

These theoretical results represent sufficient (but not necessary) conditions
for the convergence
of the forward and adjoint Monte Carlo and can be easily verified.
However, in many cases the conditions $\lVert H\rVert_{\infty}<1$ or
$\lVert H\rVert_1<1$ may be restrictive.

@@

The method for determining the transition probability has a significant impact
on the properties of the resulting approach, and in many circumstances the
choice can be the difference between convergence or divergence of the stochastic
scheme.  Two primary approaches have been considered in the literature:
\textit{uniform probabilities} and \textit{weighted probabilities}.

\subsubsection{Uniform probabilities}

With this approach, the transition matrix $P$ is such that all the non-zero elements
in each row have equal probability of occurring:
\[
\text{Forward}:\;P_{i,j}=
\begin{cases}
0 \quad \quad \quad \qquad \qquad \qquad \text{if}\quad H_{i,j}=0 \\
\frac{1}{\#(\text{non-zeros in the row)}} \quad \text{if} \quad H_{i,j}\ne 0
\end{cases}\quad
\]
\[
\text{Adjoint}:\;P_{i,j}=
\begin{cases}
0 \quad \quad \quad \qquad \qquad \qquad \text{if}\quad H_{j,i}=0 \\
\frac{1}{\#(\text{non-zeros in the column)}} \quad \text{if} \quad H_{j,i}\ne 0
\end{cases}
\]
The Monte Carlo approach resorting to this definition of the transition matrix,
in accordance to \cite{AADBTW2005}, is called \textit{Uniform Monte Carlo} (UM).

\subsubsection{Weighted probabilities}

An alternative way of defining the transition matrix consists in attributing
probabilities to the entries of the transition probability according
to the magnitude of the elements. For instance we may employ the following
definition
\[
\text{Forward}: \;
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert
H_{i,j}\rvert^p}{\sum_{k=1}^n
\lvert H_{i,k}\rvert^p}
\]
and
\[
\text{Adjoint}: \;
p(k_i=j \;\lvert\;k_{i-1}=i )=P_{i,j}=\frac{\lvert
H_{j,i}\rvert^p}{\sum_{k=1}^n
\lvert H_{k,i}\rvert^p} \,,
\]
where $p\in \mathbb{N}$.
This selection, for $p=1$, has been termed
\textit{Monte Carlo Almost Optimal} (MAO).
The reason for the ``almost optimal'' designation can been seem from
Theorem~\ref{lemma}, as the quantity
$\displaystyle \sum_{k=0}^N\frac{a_k^2}{p_k}$
is minimized when the probability vector is
defined such as $\displaystyle p_k=\frac{\lvert a_k\rvert}{\sum_{k=1}^N \lvert
a_k\rvert}$. This implies that the almost optimal probability minimizes the
$\infty$-norm of $\hat{H}$ in the forward method and the $1$-norm of $\hat{H}$
in the adjoint method. This is very important, since these two norms
provide computable upper bounds for $\rho(\hat{H})$.
In Section~\ref{sec:results}, it will be shown
that the almost optimal probability enables convergence of the MC solver,
even in cases when the employment of the uniform probability does not provide
convergence at all.

@@

As we have already seen, sufficient conditions for
the convergence of the Monte Carlo linear solvers are very restrictive (see
\cite{MASC2013}), whereas the necessary and sufficient condition requires the
computation of $\rho(H)$and $\rho(\hat{H})$, which is very expensive in general.
Many instances of matrices (perhaps with appropriate preconditioning) such
that $\rho(H)<1$ are available, but verifying $\rho(\hat{H})<1$ is generally
more problematic.  We now with to identify classes of matrices for which
the Monte Carlo linear solver is guaranteed to converge.

\subsubsection{Strictly Diagonally Dominant}
\label{sec:sdd}

@@

Introducing a MAO transition probability for the forward method we get
\[
 P_{ij}=\frac{\lvert H_{ij}\rvert}{\sum_{k=1}^n\lvert H_{ik}\rvert}
\]
therefore the entries of $\hat{H}$ are defined as follows
\[
 \hat{H}_{ij}=\frac{H^2_{ij}}{P_{ij}}=\lvert H_{ij}\rvert\bigg
(\sum_{k=1}^n\lvert
H_{ik}\rvert\bigg ).
\]
Therefore
\[
 \sum_{j=1}^n\lvert \hat{H}_{ij} \rvert = \sum_{j=1}^n \hat{H}_{ij} =\bigg
(\sum_{j=1}^n \lvert H_{ij}\rvert\bigg )\bigg (\sum_{k=1}^n\lvert
H_{ik}\rvert\bigg ) =\bigg
(\sum_{j=1}^n \lvert H_{ij}\rvert\bigg ) ^2 <1 \quad \forall i=1,\cdots, n.
\]
This automatically implies that $\rho(\hat{H})<=\lVert
\hat{H}\rVert_{\infty}<1$. Thus the forward Monte Carlo converges.
However nothing a priori can be
stated about the behavior of the adjoint method.

Instead if (\ref{sddc}) holds we can resort to a right diagonal preconditioning
and we
get an iteration matrix $H=I-AD^{-1}$ such that $\lVert H \rVert_{1}<1$.
In this case, by following a similar reasoning to the one made before, we
conclude that the adjoint method convergences thanks to $\lVert
\hat{H}\rVert_1<1$,
whereas nothing can be said in advance as concerns the forward method.

\subsubsection{Generalized Diagonally Dominant}
\label{sec:gdd}

@@

A proper subset of the generalized diagonally dominant matrices is represented
by $M$- matrices (see \cite{Ax1996}).
There are many definitions of $M$-matrices that may be found on books and all
of them are equivalent. The one presented in \cite{Saad} states the
following
\begin{defn}
A matrix $A\in\mathbb{R}^{n\times n}$ is said to be an $M$-matrix if it
satisfies the following four properties:
\begin{itemize}
 \item $a_{i,i}>0$ for $i=1,\cdots,n$
 \item $a_{i,j}\le 0$ for $i\ne j$, $i,j=1,\cdots,n$
 \item $A$ is nonsingular
 \item $A^{-1}\ge 0$
\end{itemize}
\end{defn}

A very important result about $M$-matrices enables to come up with a convergent
splitting and, therefore, a convergent fixed point scheme.
In particular Jacobi, Block Jacobi and Gauss Seidel  are
convergent
splitting for an $M$-matrix.
However, the construction of a splitting like this is not enough to ensure the
convergence of
the Monte Carlo linear solver. In fact none of these conditions is sufficient
to guarantee $\rho(\hat{H})<1$. Because of this obstacle, other options must
be considered.

@@

Here below we report the algorithm at hand

@@

This approach turns a generalized diagonally dominant matrix into a
strictly diagonally dominant matrix by rows. By substituting
$\displaystyle S_{i}=\sum^{n}_{\substack{j=1 \\ j\ne i}}\lvert a_{ij}\rvert$
at step 1 with
$\displaystyle S_{i}=\sum^{n}_{\substack{i=1 \\ i\ne j}}\lvert a_{ij}\rvert$
and by replacing
$a_{ji}=a_{ji}\cdot d_i$ with $a_{ji}=a_{ji}\cdot d_j$ we obtain the algorithm
that turns a GDDM matrix into a s.d.d. by columns.

@@

\subsubsection{Block Diagonally Dominant}
\label{sec:bdd}

In the section we consider situations in which block diagonal preconditioning
can result in a convergent Monte Carlo linear solver.
By mirroring the computations we already showed previously, the iteration
matrix $H\in\mathbb{R}^{n\times n}$ resulting from a block diagonal
preconditioning takes the form
\[
 H=I-D^{-1}A=\begin{bmatrix}0_{n_1\times n_1} & -[A_{11}]^{-1}A_{12} & \cdots &
\cdots & -[A_{11}]^{-1}A_{1p} \\
-[A_{22}]^{-1}A_{21} & 0_{n_2\times n_2} & -[A_{22}]^{-1}A_{23} &
\cdots & -[A_{22}]^{-1}A_{2p}\\
\vdots & \vdots & \ddots & \vdots & \vdots\\
\vdots & \vdots & \vdots &\ddots & \vdots \\
-[A_{pp}]^{-1}A_{p1} &  \cdots & \cdots&
-[A_{pp}]^{-1}A_{p,p-1} & 0_{n_p \times n_p}
\end{bmatrix}.
\]

For the sake of clarity, we note that the symbol "$\%$" is used in the following
subsections represents the modulus function between two integer numbers, while
the floor function is represented with the standard symbol "$\lfloor
\cdot \rfloor$".

We first consider the forward method.
By assuming that all $n_i$'s have the same value we may define
\[
 m=n_i=\text{size of a block}=\frac{n}{p}.
\]
The MAO transition probability matrix becomes:
\[
 P_{i,j}=\frac{\lvert H_{i,j} \rvert}{\sum_{k=1}^n\lvert H_{i,k}
\rvert}=\frac{\bigg \lvert \bigg ([A_{\lfloor
\frac{i}{m}\rfloor \lfloor]
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg
\rvert}{\sum_{\substack{k=1\\k\ne i}}^n\bigg \lvert \bigg
([A_{\lfloor
\frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{k}{m}\rfloor}\bigg )_{i\%m,k\%m}\bigg \rvert}
\]
Consequently, the $\hat{H}$ matrix is defined such that
\[
\hat{H}_{i,j} = \lvert H_{i,j}\rvert\bigg(\sum_{k=1}^n\lvert H_{ik}\rvert\bigg)=
\bigg \lvert \bigg ([A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg
\rvert
\sum_{\substack{k=1\\k\ne i}}^n\bigg \lvert \bigg ([A_{\lfloor
\frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{k}{m}\rfloor}\bigg )_{i\%m,k\%m}\bigg \rvert
\]
By computing the sum over a generic row of $\hat{H}$ we get:
\[
 \sum_{j=1}^n \lvert \hat{H}_{i,j}\rvert=\sum_{j=1}^n \hat{H}_{i,j} =
 \bigg ( \sum_{\substack{j=1\\j\ne i}}^n\bigg \lvert \bigg ([A_{\lfloor
\frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg \rvert \bigg ) ^2.
\]
If we focus on the norm $\lVert \hat{H}\rVert_{\infty}$, then the following
equivalence condition holds:
\[
 \lVert \hat{H}\rVert_{\infty}<1 \Leftrightarrow \sum_{\substack{j=1\\j\ne
i}}^n\bigg \lvert \bigg ([A_{\lfloor
\frac{i}{m}\rfloor \lfloor
\frac{i}{m}\rfloor}]^{-1} A_{\lfloor \frac{i}{m}\rfloor \lfloor
\frac{j}{m}\rfloor}\bigg )_{i\%m,j\%m}\bigg \rvert <1 \quad \forall
i=1,\cdots,n
\]
 A sufficient condition for this to happen is
 \begin{equation}
  \sum_{\substack{j=1\\j\ne i}}^p \lVert [A_{ii}]^{-1}A_{ij}\rVert_{\infty}<1.
    \label{block_cs}\quad \forall i=1,\ldots,p.
 \end{equation}
By defining a matrix $\tilde{H}\in \mathbb{R}^{p\times p}$ such that
\[
 \tilde{H}=\begin{bmatrix}0 & \lVert [A_{11}]^{-1}A_{12}\rVert_{\infty} &
\cdots &
\cdots & \lVert [A_{11}]^{-1}A_{1p}\rVert_{\infty} \\
\lVert [A_{22}]^{-1}A_{21}\rVert_{\infty} & 0 & \lVert
[A_{22}]^{-1}A_{23}\rVert_{\infty} &
\cdots & \lVert [A_{22}]^{-1}A_{2p}\rVert_{\infty} \\
\vdots & \vdots & \ddots & \vdots & \vdots\\
\vdots & \vdots & \vdots &\ddots & \vdots \\
\lVert [A_{pp}]^{-1}A_{p1}\rVert_{\infty} &  \cdots & \cdots&
\lVert [A_{pp}]^{-1}A_{p,p-1}\rVert_{\infty} & 0
\end{bmatrix}
\]
We can use the $\tilde{H}$ matrix just defined in order to introduce a
sufficient condition for the convergence of the forward Monet Carlo method with
a Block Diagonal preconditioning.
\begin{equation}
 \lVert \tilde{H} \rVert_{\infty}<1 \Rightarrow \lVert \hat{H}
\rVert_{\infty}<1.
\end{equation}
